---
title: "BRETIGEA_vs_CellCODE_analysis"
author: "Kailash"
date: "17/01/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Author: B P Kailash
## Purpose: Analyse the different cell deconvolution models namely BRETIGEA and CellCODE using different parameters to check which model performs best

* TRUTH data: Data with gene expression and IHC measured cell type proportions
* https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008120 
* Deconvolving the contributions of cell-type heterogeneity on cortical gene expression

Analyzing different models of BRETIGEA and CellCODE, CellCODE 80 marker gene (mg), four cell types (ct) (ast, mic, neu, oli) was chosen to be the best model - see CellCODE analysis, BRETIGEA analysis and Making the ggpairs plot sections of the code.

# Global variables ------------------------------------------------------------

```{r}
rm(list = ls())

source("scripts/utils.R")
```

# Loading libraries ------------------------------------------------------------

```{r}
.libPaths(c("libraries/", .libPaths()))
# source("scripts/libraries.R")

library(sva)
library(BRETIGEA)
library(CellCODE)
library(ggplot2)
library(GGally)
library(corrplot)
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
```

# Loading data ------------------------------------------------------------

```{r}
# 49 samples, 5 cell types, 13,469 genes
ihc_gene_exp <- read.csv(paste("data/","cortical_gene_expression.csv", sep = ""),
                         stringsAsFactors = FALSE)
row.names(ihc_gene_exp) <- ihc_gene_exp$X
ihc_gene_exp <- subset(ihc_gene_exp, select = -X)

# 49 samples, 6 cell types
ihc_cell_freq <- read.csv(paste("data/","cortical_cell_frequencies.csv", sep = ""),
                          stringsAsFactors = FALSE)
row.names(ihc_cell_freq) <- ihc_cell_freq$X
ihc_cell_freq <- subset(ihc_cell_freq, select = -X)
colnames(ihc_cell_freq) <- c("ast", "end", "oli", "mic", "neu")

# Checking if the samples are in the same order
stopifnot(row.names(ihc_gene_exp) == row.names(ihc_cell_freq))

# Choosing 4 cell types
ihc_cell_freq <- subset(ihc_cell_freq, select = c(ast,mic,neu,oli))
cell_types <- colnames(ihc_cell_freq)

mdata <- read.table(paste("data/", "ihc_pathoVariables.txt", sep = ""),
                    stringsAsFactors = FALSE, sep = "\t")

ihc_projids <- as.numeric(sapply(strsplit(row.names(ihc_gene_exp), "X"), function(x){x[2]}))
row.names(ihc_gene_exp) <- row.names(ihc_gene_exp) <- ihc_projids
ihc_mdata <- mdata[match(ihc_projids, mdata$projid),]
```

# CellCODE analysis -------------------------------------------------------

## WE_m = UDV_t

* marker genes has Marker and its associated cell type 
* (markers_df_human_brain, colnames = c("markers", "cell"))
* cellTypes - list of cell types for making the W matrix 
* (unique(markers_df_human_brain[,2]))
* numberMarkersPerCelltype - list of number of markers to take from each cell type 
* c(100,100,100,100,100,50)

```{r}
make_w_matrix <- function(markerGenes, cellTypes, numberMarkersPerCellType){
  mg <- c()
  for (i in 1:length(cellTypes)){
    mg_ct <- head(markerGenes$markers[which(markerGenes$cell == cellTypes[i])], n = numberMarkersPerCellType[i])
    mg <- c(mg, mg_ct)
  }
  
  w <- matrix(0L, nrow = sum(numberMarkersPerCellType), ncol = length(cellTypes))
  colnames(w) <- cellTypes
  rownames(w) <- mg
  
  sum_number_markers <- rep(0, length(numberMarkersPerCellType))
  for (i in 1:length(sum_number_markers)){
    if (i == 1){
      sum_number_markers[i] = numberMarkersPerCellType[i]
    } else {
      sum_number_markers[i] = numberMarkersPerCellType[i] + sum_number_markers[i-1]
    }
  }
  
  for (i in 1:length(cellTypes)){
    if (i == 1){
      w[,cellTypes[i]][1:numberMarkersPerCellType[i]] = rep(1, numberMarkersPerCellType[i])
    } 
    else{
      w[,cellTypes[i]][(sum_number_markers[i-1]+1) : sum_number_markers[i]] = rep(1, numberMarkersPerCellType[i])
    }
  }
  
  return(list(w, mg))
}

get_SPV_BrainRegion <- function(BrainRegion, MarkerGenes){
  # Getting the expression matrices
  BR <- readBMfile(BrainRegion)
  colnames(BR) <- genes_mapped
  BR$sample_gp <- find_sample_group(BrainRegion)
  
  BR_AD_Ctl <- BR[c(which(BR$sample_gp == "Control"), which(BR$sample_gp == "Definite AD")),]
  sample_gp <- BR_AD_Ctl$sample_gp
  BR_AD_Ctl <- subset(BR_AD_Ctl, select = -c(sample_gp))
  
  mg_new <- intersect(MarkerGenes, colnames(BR_AD_Ctl))
  w_new <- w[mg_new,]
  
  # Finding E_m
  # NOTE: length(intersect(mg, colnames(bm36))) = 4181 != 5500
  # NOTE: length(which(is.na(colnames(bm36) == ""))) = 1029
  
  E_m <- t(BR_AD_Ctl[,mg_new])
  colnames(E_m) <- rownames(BR_AD_Ctl)
  spvs <- as.data.frame(my_getAllSPVs(brainRegion = paste(BrainRegion, "IHC", sep = ""), data = E_m, grp = sample_gp, dataTag = w_new,
                                      method = "mixed", plot = T, mix.par = 0.3))
}

get_SPV_ihc <- function(GeneExpression, MG){
  ad_ids <- ihc_mdata$projid[which(ihc_mdata$ceradsc %in% c(3,4))]
  ctl_ids <- ihc_mdata$projid[which(ihc_mdata$ceradsc %in% c(1,2))]
  sample_gp <- rep(1, length(row.names(GeneExpression)))
  sample_gp[which(row.names(GeneExpression) %in% ad_ids)] = "Definite AD"
  sample_gp[which(row.names(GeneExpression) %in% ctl_ids)] = "Control"
  
  mg_new <- intersect(MG, colnames(GeneExpression))
  w_new <- w[mg_new,]
  E_m <- t(GeneExpression[,mg_new])
  colnames(E_m) <- rownames(GeneExpression)
  spvs <- as.data.frame(my_getAllSPVs(brainRegion = "IHC", 
                                      data = E_m, grp = sample_gp, dataTag = w_new,
                                      method = "mixed", plot = T, mix.par = 0.3))
}
```

# BRETIGEA analysis -------------------------------------------------------

```{r}
find_brainCells_BR <- function(BrainRegion, nMarker){
  # Getting the expression matrices
  BR <- readBMfile(BrainRegion)
  colnames(BR) <- genes_mapped
  BR$sample_gp <- find_sample_group(BrainRegion)
  BR_AD_Ctl <- BR[c(which(BR$sample_gp == "Control"), which(BR$sample_gp == "Definite AD")),]
  sample_gp <- BR_AD_Ctl$sample_gp
  BR_AD_Ctl <- t(subset(BR_AD_Ctl, select = -c(sample_gp)))
  return(as.data.frame(brainCells(BR_AD_Ctl, nMarker, species = "human", celltypes = c("ast", "mic", "neu", "oli"))))
}

# brainCells is a BRETIGEA function
find_brainCells_ihc <- function(GeneExpression, nMarker){
  return(as.data.frame(brainCells(t(GeneExpression), nMarker, species = "human")))
}
```

# Making the ggpairs plot -------------------------------------------------

## All the following plots will be for 4 cell-types, but different mg sets

```{r}
bretigea_10 <- find_brainCells_ihc(ihc_gene_exp, 10)
bretigea_20 <- find_brainCells_ihc(ihc_gene_exp, 20)
bretigea_50 <- find_brainCells_ihc(ihc_gene_exp, 50)

ip <- make_w_matrix(markers_df_human_brain, c("ast", "mic", "neu", "oli"), c(10,10,10,10))
w <- ip[[1]]
mg <- ip[[2]]
cellcode_10 <- get_SPV_ihc(ihc_gene_exp, MG = mg)

ip <- make_w_matrix(markers_df_human_brain, c("ast", "mic", "neu", "oli"), c(20,20,20,20))
w <- ip[[1]]
mg <- ip[[2]]
cellcode_20 <- get_SPV_ihc(ihc_gene_exp, MG = mg)

ip <- make_w_matrix(markers_df_human_brain, c("ast", "mic", "neu", "oli"), c(50,50,50,50))
w <- ip[[1]]
mg <- ip[[2]]
cellcode_50 <- get_SPV_ihc(ihc_gene_exp, MG = mg)
```

# cell-type can be "ast", "mic", "neu", and "oli"

```{r}
make_ggpairs_prediction_plot <- function(cell_type){
  d1 <- bretigea_10[, cell_type]
  d2 <- cellcode_10[, cell_type]
  d3 <- bretigea_20[, cell_type]
  d4 <- cellcode_20[, cell_type]
  d5 <- bretigea_50[, cell_type]
  d6 <- cellcode_50[, cell_type]
  d7 <- ihc_cell_freq[, cell_type]
  
  data <- as.data.frame(cbind(d1, d2, d3, d4, d5, d6, d7))
  colnames(data) <- c("BRETIGEA 40mg", "CellCODE 40mg", "BRETIGEA 80mg", "CellCODE 80mg", "BRETIGEA 200mg", "CellCODE 200mg", "Truth")
  
  (pair_plot <- ggpairs(data, upper = list(continuous = wrap("cor", size = 7)),
                       title = paste("Predictions for ", cell_type, " cell type", sep = "")))
  
  ggsave(path = "plots/",
         plot = pair_plot,
         device = "pdf",
         filename = paste0(cell_type, "_analysisDeconvolutionAlgos.pdf"),
         width = 15, height = 12, dpi = 300)
}

make_ggpairs_prediction_plot("ast")
make_ggpairs_prediction_plot("mic")
make_ggpairs_prediction_plot("neu")
make_ggpairs_prediction_plot("oli")
```

# Comparing CellCODE vs BRETIGEA ------------------------------------------

```{r}
BM10_findBrainCells <- find_brainCells_BR(BrainRegion = "BM_10", nMarker = 20)

ip <- make_w_matrix(markers_df_human_brain, c("ast", "mic", "neu", "oli"), c(20,20,20,20))
w <- ip[[1]]
mg <- ip[[2]]
BM10_spv <- get_SPV_BrainRegion("BM_10", mg)

# Correlation matrix 
BM10_findBrainCells <- subset(BM10_findBrainCells, select = c(ast, mic, neu, oli))
cor_matrix <- cor(BM10_spv, BM10_findBrainCells)
corrplot(cor_matrix,
         type = "upper",
         col = brewer.pal(n = 8, name = "YlOrRd"),
         order = "hclust",
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "gray95",
         number.cex = 0.75,
         mar = c(4,0,4,0),
         title = "BM10: BRETIGEA 80mg vs CellCODE 80mg",
         outline = T,
         addgrid.col = "darkgray",
         rect.lwd = 5,
         addrect = 4,
         rect.col = "black"
)
```

# Scatter plot for Neuronal cell type proportion estimates

```{r}
neu_proportions <- data.frame(cbind(BM10_spv$neu, BM10_findBrainCells$neu))
colnames(neu_proportions) <- c("CellCODE", "BRETIGEA")

scatter_neu <- ggplot(neu_proportions, aes(x = BRETIGEA,y = CellCODE)) + 
  geom_point(size = 2,color = "orange") + theme_bw(base_size = 20)  + 
  theme(axis.text = element_text(size = 15, face = "bold"),
        axis.title = element_text(size = 15,face = "bold"), 
        plot.title = element_text(size = 18, face = "bold")) +
  geom_smooth(method ='lm',color = "red") + stat_cor(method = "pearson") +
  ggtitle("BM10: Scatter plot of estimated neuronal cell type frequencies")

scatter_neu
```

# Scatter plot between CellCODE 80mg and True frequencies

```{r}
neu_proportions <- data.frame(cbind(d4, d7))
colnames(neu_proportions) <- c("CellCODE", "True")

scatter_neu <- ggplot(neu_proportions, aes(x = CellCODE,y = True)) + 
  geom_point(size = 2,color = "orange") + theme_bw(base_size = 20)  + 
  theme(axis.text = element_text(size = 15, face = "bold"),
        axis.title = element_text(size = 15,face = "bold"), 
        plot.title = element_text(size = 18, face = "bold")) +
  geom_smooth(method ='lm',color = "red") + stat_cor(method = "pearson") +
  ggtitle("BM10: Scatter plot of estimated neuronal cell type frequencies")

scatter_neu
```
